---
title: "02_Data Munging"
author: "Brandon Foster, Ph.D."
output: html_notebook
---

This script should walk the user through all of the steps necessary to clean and transform the final data frames for any analyses. 



First, let's pull out the items for the two measures of interest, the SEP items and the science interest items, into their own respective data frames. 
```{r, warning=FALSE, message=FALSE, tidy=TRUE}
# load packages
if (!require("pacman")) install.packages("pacman")
pacman::p_load(stringr, rmarkdown)

# run the 01 script
render("../rmds/01_Import.Rmd", "all", output_dir = "reports")

# Join the two data frames ----
joined.dat <- pre.dat %>%
  inner_join(post.dat)
```

Create a vector that represents school mandated participation in the science fair. 
```{r, munge.1, warning=FALSE, message=FALSE, tidy=TRUE}
# Create vector to represent school mandated voluntary participation ----
joined.dat <- joined.dat %>% mutate(participateReq = ifelse(SchoolID == 224 | SchoolID == 225 | SchoolID == 227 | SchoolID == 229 | SchoolID == 321 | SchoolID == 423, 'mandated', 'voluntary'))
```

Create a vector that represents whether the student provided a title for the science fair. Note that when the string responses were examined, some students indicated in some way that they either "didn't know" the name of the project or "didn't" have to have one. These string responses are set to missing. 
```{r, munge.2, warning=FALSE, message=FALSE, tidy=TRUE}
#  Create vector to represent whether student provided name of science fair project ----
joined.dat <- joined.dat%>% mutate(
    providedTitle = ifelse(str_detect(post44_SFtitle, c("NR", "999", "IDK", "did not have one", "Don't 
                                                        know", "We didn't have one", "Don't Remember", 
                                                        "I can't remember", "I didn't do it", "I don't 
                                                        remember", "[no response]", " ")), 
                           "not provided", "provided"))

# Munge the post38_SFparticipate variable ----
# trim white spaces
joined.dat$post38_SFparticipate <- gsub("(^\\s+)|(\\s+$)", "", 
                                        joined.dat$post38_SFparticipate)
joined.dat$post38_SFparticipate[joined.dat$post38_SFparticipate=="yes"] <- "Yes"
joined.dat$post38_SFparticipate[joined.dat$post38_SFparticipate=="no"] <- "No"
joined.dat$post38_SFparticipate[joined.dat$post38_SFparticipate==""] <- "Not Reported"
```

Create seperate RDS files for the SEP and science interest items.
```{r, warning=FALSE, message=FALSE, tidy=TRUE}
# Select the SEP items ----
sep.dat <-select(joined.dat, SchoolID, StudentID, contains("preSEP_"), contains("postSEP_"), participateReq, post38_SFparticipate, post44_SFtitle, providedTitle)

# Select the opinions about science items ----
opsci.dat <-select(joined.dat, SchoolID, StudentID, contains("preInt_"), contains("postInt_"), participateReq, post38_SFparticipate, post44_SFtitle, providedTitle)

# Save the tidy-ed data ----
saveRDS(joined.dat, file = "../data/joined.dat")
saveRDS(sep.dat, file = "../data/sep.dat")
saveRDS(opsci.dat, file = "../data/opsci.dat")
```


