---
title: "03_Analyses"
output: html_notebook
---

This script should walk the user through any meaningful chunk of analyses. I like to think of these scripts as analgous to chapters in a book- if a set of analyses warrants its own chapter, then it should have its own notebook. For example, an obvious way to break up a simple regression analysis up would be to create seperate notebooks for: descriptives, EDA, and the regression. You might also find yourself conducting several large scale analyses. In this case I would adopt a numeric naming system as follows: 

- 03_01_Descriptives

- 03_02_EDA

- 03_03_Regression

- 04_01_Descriptives

- 04_02_EDA

- 04_03_LCA

 ```{r setup, include=FALSE}
 knitr::opts_knit$set(root.dir = 'rmds')
 ```
 
```{r}
# Initialize an empty list to store the results ----
results <- list()

# Load the transformed data ----
readRDS("data/transformed.rds") -> df

# Specific transformations ----

# Models ----

# Tables ----

# Plots ----

# Save the results object ----
saveRDS(results, file = "data/results.rds")
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file).
