---
title: "EFA Analyses of the Science Interest Measure"
author: "Brandon Foster, Ph.D."
date: "`r format(Sys.time(), '%d %B, %Y, %H:%M')`"
output:
  html_document:
  html_notebook: null
---

# Introduction

The following report outlines exploratory analyses to ascertain the dimensional structure of the Science Interest measure. In the following analyses, sample of 447 students responded to a measure of science interest measure, with items measuring their: desire to do science, career interest in science, and self-concept in science. Confirmatory analyses for a three dimensional structure of the measure failed to produce an adequate fit. As such, in order to determine whether distinctions among the items are justified, an exploratory factor analysis was conducted. 

## Import the data 
Let's bring in the data for analyses, load the necessary packages, load any functions necessary for the analyses, and subset out the items so that they can be used in the subsequent analyses. Please note, that I have chosen to use all of the respondents to the survey, which was administered during the post-science fair exposure period. This is slightly different than the sample we will use to investigate outcomes. I do not expect the factor solutions would look that different if those few students who didn't participate in the science fair were removed from the analyses.

```{r, import, message=FALSE,  error=FALSE, warning=FALSE, tidy=TRUE}
# load packaes
if (!require("pacman")) install.packages("pacman")
pacman::p_load(MplusAutomation, psych, tidyr, foreign, ggplot2, ggalt, ggthemes, readr, dplyr, knitr, scales, pander, lavaan, kableExtra, GPArotation, nFactors, stringr)

# Import the data ----
joined.dat <- readRDS(file = "../data/joined.dat")

# Load the custom functions ----
source("../functions/functions.r")

# Document options ----
panderOptions('digits', 2)
panderOptions('round', 2)
panderOptions('keep.trailing.zeros', TRUE)

# Munge data ----
items <- joined.dat %>%
  dplyr::select(s_preInt_Des_15, s_preInt_Des_17, s_preInt_Des_21, s_preInt_Des_24, s_preInt_Des_26, 
         s_preInt_Des_29, s_preInt_Des_33, s_preInt_Car_16, s_preInt_Car_18, s_preInt_Car_20, 
         s_preInt_Car_23, s_preInt_Car_25, s_preInt_Car_28, s_preInt_Car_30, s_preInt_Car_32, 
         s_preInt_Car_34, s_preInt_Car_36, s_preInt_Self_19, s_preInt_Self_22, 
         s_preInt_Self_27, s_preInt_Self_31 ,s_preInt_Self_35)

# munge the variable names to remove pre_ and post_ prefixes
items.stringr.prune <- items %>% 
  rename_(.dots=setNames(names(.), gsub("s_preInt_", "", names(.))))
```

# EFA: Run #1

The following analyses uses an oblique rotation and a principal axis method for extraction. This will produce an [OLS solution which is not as sensitive to improper matrices as is the maximum likelihood method, and will sometimes produce more interpretable results](http://personality-project.org/r/psych/HowTo/factor.pdf). The following analysis utilizes a promax rotation, which is standard for correlated factors. 
```{r, efa.1, message=FALSE,  error=FALSE, warning=FALSE, tidy=TRUE}
# run the EFA model for each numer of factors. 
efa.2.5.promax <- lapply(2:5, function(x) {fa(items, nfactors=x, fm="pa", rotate = "promax", 
                                              max.iter = 500)})
```

Next, the fit statistics for all four factor models are provided. An easy way to ascertain the best "fitting" model is to look for the model with the lowest BIC. The TLI and RMSEA values can then be used to examine how well that model fit the data. Results indicate that the 2-factor model exhibits the smallest BIC statistic, followed by the 5-factor solution. Of these two models, the 5-factor model shows the smallest fit statistics with RMSEA value = about .05, and TLI > .90. This result indicates that the patterns in the factor loadings for both should be examined. 

```{r, efa.1.fit, message=FALSE,  error=FALSE, warning=FALSE, tidy=TRUE}
tibble(
  "Number of Factors" = c(2, 3, 4, 5),
  "Tucker-Lewis Index (TLI)" = c(efa.2.5.promax[[1]]$TLI, efa.2.5.promax[[2]]$TLI,
                                 efa.2.5.promax[[3]]$TLI, efa.2.5.promax[[4]]$TLI), 
  "RMSEA" = c(efa.2.5.promax[[1]]$RMSEA[1], efa.2.5.promax[[2]]$RMSEA[1], 
              efa.2.5.promax[[3]]$RMSEA[1], efa.2.5.promax[[4]]$RMSEA[1]),
  "Lower 90% RMSEA" = c(efa.2.5.promax[[1]]$RMSEA[2], efa.2.5.promax[[2]]$RMSEA[2], 
                        efa.2.5.promax[[3]]$RMSEA[2], efa.2.5.promax[[4]]$RMSEA[2]),
  "Upper 90% RMSEA" = c(efa.2.5.promax[[1]]$RMSEA[3], efa.2.5.promax[[2]]$RMSEA[3], 
                       efa.2.5.promax[[3]]$RMSEA[3], efa.2.5.promax[[4]]$RMSEA[3]),
  "Bayesian Information Criterion (BIC)" = c(efa.2.5.promax[[1]]$BIC, 
                                             efa.2.5.promax[[2]]$BIC, 
                                             efa.2.5.promax[[3]]$BIC, 
                                             efa.2.5.promax[[4]]$BIC)
  ) %>%
  kable(digits = 3, format="html", caption="Fit statistics for the four different EFA 
          models") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## 2-factor solution
Now we will take a look at the details for the 2-factor solution. First, let's look at a factor diagram for the 2-factor solution. This can be an easy way to visualize the solution. 

```{r, efa.1.2diag, message=FALSE,  error=FALSE, warning=FALSE, tidy=TRUE, results='asis'}
fa.diagram(efa.2.5.promax[[1]], main="Factor diagram for the 2-factor solution")
```

From a substantive perspective, this solutions is quite intuitive. When inspecting the diagram you can see that the first factor loads onto all of the career interest in science items **and** the desire to do science items. We might expect that the items for these two concepts are highly correlated with one another... that students who have a high desire to do science would most likely indicate that they were interested in a career in science. Said in other words, we would expect the responses to these items to covary. The fact that the first factor loads onto all of these items means that they likely do not differentiate distinctly enough to warrant creating a factor score for each. The second factor captures all of the self-concept items. Further, the factor loadings for the indicators for this factor are all >= .50. In addition, the correlation between the first and second factor is = .60, squaring this value we find that the the shared variance between the factors is 16%, so while there is a relationship between the two factors, each is measuring something distinct. Overall, this is about as clean cut a factor solution as you can find in an exploratory analyses. However, please note that while this solution "makes sense," we still don't have an understanding of how well the items are measuring each factor. We know from examining the RMSEA and TLI that those values fall in the range of what could be described as marginally acceptable. We'll need to dig deeper to to understand how well each item is measuring the respective construct. 

The more specific information in the output provides additional insights.

```{r, efa.1.2out, message=FALSE,  error=FALSE, warning=FALSE, tidy=TRUE}
# print the 2-factor solution
print(fa(items, nfactors=2, fm="pa", rotate = "promax", 
         max.iter = 500), digits=2, cut=.3, sort=TRUE)

# print a plot with scree and parallel tests
ev <- eigen(cor(items, use="complete.obs")) # get eigenvalues
ap <- parallel(subject=nrow(items),var=ncol(items),
  rep=100,cent=.05)
nS <- nScree(x=ev$values, aparallel=ap$eigen$qevpea)
plotnScree(nS)
```

EFA: What is the structure of the latent factors that produced the covariances among the observed indicators (factor = predictor)?


Rotations
The idea of rotation is to redefine the factor loadings to obtain simple structure
Ø Each factor should have indicators with strong loadings § Obvious which indicators measure it (+/-) and which don’t
Ø Each indicator should load strongly on only one factor
§ Know what each item is ‘for’
§ Construct measured is readily identifiable
§ Indicators should have large communalities

Promax rotation:
Computes best orthogonal solution and then “relaxes” orthogonality constraints to better “spear” indicator clusters with factor vectors (give simpler structure)

### Interpreting the solution
Interpretation is the process of “naming factors” based on the indicators that “load on” them. Which indicators “load” is decided based on a “cutoff”
Ø Cutoffs usually range from .3 to .4 ( +/- )
Ø Note that significance tests of loadings are not usually given!!
§ Although can be obtained separately though other procedures or in Mplus

However, there is a tradeoff, as setting a higher criteria for factor loadings might leave some factors ill-defined or completely undefined, while setting a lower criteria increases the likelihood that indicators load onto more than one factor.

# ICM-CFA 

The models that follow explore theoretical fit of models using confirmatory factor analyses. Based on theory, we might exepect that a three factor model best fits the data. However, the EFA analyses above indicate that a 2 factor model might be most appropriate. The code that follows will establish if either of these models fit the data or if additional modifications are needed to establish a better fit. 

## ICM-CFA approach 3-factor 

The following analysis fits a traditional CFA model to the science interest items. The specifications of the model fit with theoretical conceptualizations of the measure, as it assess the fit of a model that contains factors for: desire to do science, career interest in science, and self-interest in science. Analyses are conducted in MPlus, whereby items are considered as continuous and the restricted maximum likelihood estimation procedure is utilized. Code for the models are generated in R, and the MPlusAutomation package is used to compile code in MPlus. 

```{r, ICM_CFA_3.cont, message=FALSE,  error=FALSE, warning=FALSE, tidy=TRUE}
# construct the MPlus code for the model
icm.cfaCont.3 <- mplusObject(
  TITLE = "Title: EFA of 3-factor solution using the ICM-CFA approach considering indicators 
  as continuous;",
  VARIABLE = "
  Missing are all (-99, -9, -8, -1);,
  USEVARIABLES =    Des_15  Des_17  Des_21  
                    Des_24 Des_26  Des_29  
                    Des_33  Car_16 Car_18  
                    Car_20  Car_23  Car_25 
                    Car_28  Car_30  Car_32  
                    Car_34 Car_36  Self_19 
                    Self_22 Self_27 Self_31 
                    Self_35;",
  ANALYSIS = "
  ITERATIONS = 10000;
  ESTIMATOR = MLR;
  ! MLR = maximum likelihood estimation with robust standard errors is used",
  MODEL = "
  DES BY Des_15  Des_17  Des_21  Des_24 Des_26  Des_29  Des_33;  
  CAREER BY Car_16 Car_18  Car_20  Car_23  Car_25 Car_28  Car_30
  Car_32  Car_34 Car_36;  
  SELF BY Self_19 Self_22 Self_27
  Self_31 Self_35;
  ! ICM-CFA models for the latent variables in the analysis
  ! (measurement component of the model)",
  OUTPUT = "STDYX CINTERVAL MODINDICES(3.8);
  ! STDYX standardizes parameters with the variances of the continuous latent variables and 
  ! the variances of the background and outcome variables
  ! CINTERVAL option is used to request confidence intervals for parameter estimates
  ! wherein SYMMETRIC intervals is the default for frequentist estimation (such as MLR),
  ! producing 90%, 95% and 99% symmetric confidence intervals by default in program output",
rdata = items.stringr.prune)

# compile the code and send to mplus to run the model
icm.cfaCont.3_mplus <- mplusModeler(icm.cfaCont.3,
  dataout = c("../data/06_Analayses_icm.cfaCont.3.csv"),
  modelout = c("../mplus/06_Analayses_icm.cfaCont.3.inp"), run=TRUE)

# extract model summaries
icm.cfaCont.3_summary <- readModels("../mplus/06_Analayses_icm.cfaCont.3.out")
```

In general, the model exhibits inadequate fit. The RMSEA value, and its associated 90% confidence intervals are above the .05 threshold, the CFI and TLI are below .90. The SRMR value falls within a model that exhibits acceptable fit with a value < .80. However, because the other values are not in line with an acceptable fitting model, we can not say the theoretical model fit the data. 

```{r, ICM_CFA_3.cont.sum, message=FALSE,  error=FALSE, warning=FALSE, tidy=TRUE}
# extract model summary statistics
icm.cfaCont.3_summaryStats <- dplyr::select(icm.cfaCont.3_summary$summaries, CFI, TLI, RMSEA_Estimate, RMSEA_90CI_LB, RMSEA_90CI_UB, SRMR, AIC, AICC, BIC, aBIC, AICC) 

# print table for the summary statistics
icm.cfaCont.3_summaryStats %>%
  kable(digits = 3, format="html", caption="Model fit statistics for the 3-factor ICM-CFA 
        model with contintuous indicators") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```
The table below provides the parameter estimates for the model. 
```{r, ICM_CFA_3.cont.param, message=FALSE,  error=FALSE, warning=FALSE, tidy=TRUE, results='markup'}
# print model parameters
icm.cfaCont.3_summary$parameters 
```

Modification indices shows the presence of items that, if freed in the model, several items would load onto factors different than anticipated from the theoretically predicated model.  
```{r, ICM_CFA_3.cont.mi, message=FALSE,  error=FALSE, warning=FALSE, tidy=TRUE}
# print model modification indices
icm.cfaCont.3_summary$mod_indices %>%
  filter(operator=="BY") %>%
  kable(digits = 3, format="html", caption="Model modification indices for the 3-factor 
        ICM-CFA model with contintuous indicators") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## ICM-CFA approach 2-factor 

The following analysis fits a traditional CFA model to the science interest items. The specifications of the model fit with the results for a 2-factor structure that emerged from the EFA analyses with factors for: desire to do science/career interest in science and self-interest in science. Analyses are conducted in MPlus, whereby items are considered as continuous and the restricted maximum likelihood estimation procedure is utilized. 

```{r, ICM_CFA_2.cont, message=FALSE,  error=FALSE, warning=FALSE, tidy=TRUE}
# construct the MPlus code for the model
icm.cfaCont.2 <- mplusObject(
  TITLE = "Title: EFA of 2-factor solution using the ICM-CFA approach considering indicators 
  as continuous;",
  VARIABLE = "
  Missing are all (-99, -9, -8, -1);,
  USEVARIABLES =    Des_15  Des_17  Des_21  
                    Des_24 Des_26  Des_29  
                    Des_33  Car_16 Car_18  
                    Car_20  Car_23  Car_25 
                    Car_28  Car_30  Car_32  
                    Car_34 Car_36  Self_19 
                    Self_22 Self_27 Self_31 
                    Self_35;",
  ANALYSIS = "
  ITERATIONS = 10000;
  ESTIMATOR = MLR;
  ! MLR = maximum likelihood estimation with robust standard errors is used",
  MODEL = "
  DESCAREER BY Des_15  Des_17  Des_21  Des_24 Des_26  Des_29  Des_33
  Car_16 Car_18  Car_20  Car_23  Car_25 Car_28  Car_30
  Car_32  Car_34 Car_36;  
  SELF BY Self_19 Self_22 Self_27
  Self_31 Self_35;
  ! ICM-CFA models for the latent variables in the analysis
  ! (measurement component of the model)",
  OUTPUT = "STDYX CINTERVAL MODINDICES(3.8);
  ! STDYX standardizes parameters with the variances of the continuous latent variables and 
  ! the variances of the background and outcome variables
  ! CINTERVAL option is used to request confidence intervals for parameter estimates
  ! wherein SYMMETRIC intervals is the default for frequentist estimation (such as MLR),
  ! producing 90%, 95% and 99% symmetric confidence intervals by default in program output",
rdata = items.stringr.prune)

# compile the code and send to mplus to run the model
icm.cfaCont.2_mplus <- mplusModeler(icm.cfaCont.2,
  dataout = c("../data/06_Analayses_icm.cfaCont.2.csv"),
  modelout = c("../mplus/06_Analayses_icm.cfaCont.2.inp"), run=TRUE)

# extract model summaries
icm.cfaCont.2_summary <- readModels("../mplus/06_Analayses_icm.cfaCont.2.out")
```

In general, the model exhibits inadequate fit. The RMSEA value, and its associated 90% confidence intervals are above the .05 threshold, the CFI and TLI are below .90. The SRMR value falls within a model that exhibits acceptable fit with a value < .80. However, because the other values are not in line with an acceptable fitting model, we can not say the theoretical model fit the data. 

```{r, ICM_CFA_2.cont.sum, message=FALSE,  error=FALSE, warning=FALSE, tidy=TRUE}
# extract model summary statistics
icm.cfaCont.2_summaryStats <- dplyr::select(icm.cfaCont.2_summary$summaries, CFI, TLI, RMSEA_Estimate, RMSEA_90CI_LB, RMSEA_90CI_UB, SRMR, AIC, AICC, BIC, aBIC, AICC) 

# print table for the summary statistics
icm.cfaCont.2_summaryStats %>%
  kable(digits = 3, format="html", caption="Model fit statistics for the 3-factor ICM-CFA 
        model with contintuous indicators") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```
The table below provides the parameter estimates for the model. 
```{r, ICM_CFA_2.cont.param, message=FALSE,  error=FALSE, warning=FALSE, tidy=TRUE}
# print model parameters
icm.cfaCont.2_summary$parameters 
```

Again, modification indices show factors pulling on items that we would not expect are associated with the factors. Clearly, there is something within these items that are associated with other factors. Ignoring these cross-loadings in a model would lead to mispecification of the model. Enter the EFA ESEM models, next. 

```{r, ICM_CFA_2.cont.mi, message=FALSE,  error=FALSE, warning=FALSE, tidy=TRUE}
# print model modification indices
icm.cfaCont.2_summary$mod_indices %>%
  filter(operator=="BY") %>%
  kable(digits = 3, format="html", caption="Model modification indices for the 3-factor 
        ICM-CFA model with contintuous indicators") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

# EFA ESEM

Why pursue an EFA within an exploratory structural equation modeling framework? First, ESEM provides an overarching framework allowing for the combination of CFA, EFA, and SEM into a single model. ESEM thus incorporates the benefits from each technique into a single analytic framework. In moving from an EFA to CFA model, a major assumption that we make about our indicators is that they load independently onto only 1 factor. Indicators that cross-load onto other factors are generally frowned upon. Estimates of model fit, as well as correlations between factors can be over estimated using the traditional approach of moving from an EFA to a CFA solution. As such, historically, within applied circles, the inclusion of cross-loading indicators into a model substantively changes the meaning of the factors. 

A perhaps more critical issue is whether the factor itself is adequately captured, from a psychometric perspective, as being primarily reflected in its a priori indicators. Indeed, whenever results show large and hard to explain cross-loadings suggesting that some specific factors are mainly reflected in unexpected items rather than in their a priori items, then alternatives models should be explored.
 
Forcing cross-loadings to be exactly zero involves ignoring some potentially true influence of a factor on indicators presenting some residual association with these factors over and above their association with their main a priori factor.

ESEM EFA 2-factor
```{r, efaSEM2.cont, message=FALSE,  error=FALSE, warning=FALSE, tidy=TRUE}
# construct the MPlus code for the model
efa.esem.2.obj.cont <- mplusObject(
  TITLE = "Title: EFA of 2-factor solution using ESEM considering indicators as continuous;",
  VARIABLE = "
  Missing are all (-99, -9, -8, -1);,
  USEVARIABLES =    Des_15  Des_17  Des_21  
                    Des_24 Des_26  Des_29  
                    Des_33  Car_16 Car_18  
                    Car_20  Car_23  Car_25 
                    Car_28  Car_30  Car_32  
                    Car_34 Car_36  Self_19 
                    Self_22 Self_27 Self_31 
                    Self_35;",
  ANALYSIS = "
  ITERATIONS = 10000;
  ESTIMATOR = MLR;",
  MODEL = "
  f1-f2 BY Des_15  Des_17  Des_21  Des_24 
  Des_26  Des_29  Des_33  Car_16
  Car_18  Car_20  Car_23  Car_25
  Car_28  Car_30  Car_32  Car_34 
  Car_36  Self_19 Self_22 Self_27
  Self_31 Self_35 (*1);",
  OUTPUT = "SAMPSTAT MODINDICES(3.8) STANDARDIZED tech1;",
rdata = items.stringr.prune)

# examine the mplus model syntax
#str(pathmodel, max.level = 1, nchar.max = 30)

# compile the code and send to mplus to run the model
efaSEM2.cont_mplus <- mplusModeler(efa.esem.2.obj.cont,
  dataout = c("../data/06_Analayses_EfaSEMCont.2.csv"),
  modelout = c("../mplus/06_Analayses_EfaSEMCont.2.inp"), run=TRUE)

# extract model summaries
efaSEM2.cont_summary <- readModels("../mplus/06_Analayses_EfaSEMCont.2.out")

# extract model summary statistics
efaSEM2.cont_summaryStats <- dplyr::select(efaSEM2.cont_summary$summaries, CFI, TLI, RMSEA_Estimate, RMSEA_90CI_LB, RMSEA_90CI_UB, SRMR, AIC, AICC, BIC, aBIC, AICC) 

# model parameters
efaSEM2.cont_summary$parameters 

# model modification indices
efaSEM2.cont_summary$mod_indices %>%
  kable(digits = 3, format="html", caption="Model modification indices for the 2-factor EFA ESEM") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```
