---
title: "SEP IRT (Pre-Administration Period)"
author: "Brandon Foster, Ph.D."
date: "`r format(Sys.time(), '%d %B, %Y, %H:%M')`"
output:
  html_document:
    theme: simplex
    toc: TRUE
---

# Data Screeining 

First, let's bring in the data and munge it appropriately. 
```{r, import, message=FALSE, warning=FALSE, tidy=TRUE}
# Set working directory ----
#setwd("/Users/bfoster/Desktop/2017-edc/science-fairs-analyses")
# Load packaes ----
if (!require("pacman")) install.packages("pacman")
pacman::p_load(psych, ggplot2, ggalt, ggthemes, 
    readr, dplyr, knitr, scales, pander, kableExtra, stringr, scales,
    mirt, ltm, tidyverse, formattable, gridExtra, tidyverse, broom)

# Import the data ----
joined.dat <- readRDS(file = "../data/joined.dat")

# Munge data ----
items <- joined.dat %>%
  dplyr::select(StudentID, s_preSEP_01, s_preSEP_02, s_preSEP_03, 
                s_preSEP_04, s_preSEP_05, s_preSEP_06, 
                s_preSEP_07, s_preSEP_08, s_preSEP_09, 
                s_preSEP_10, s_preSEP_11, s_preSEP_12, 
                s_preSEP_13, s_preSEP_14) %>%
  rename_(.dots=setNames(names(.), gsub("s_postInt_", "", names(.))))

# create dataframe for item reference
item.ref <- tibble(
  Item = colnames(items)[-1],
  Number = 1:14)
```

# Full measure

## Item Descriptives

The syntax below creates the item statistics using the `ltm` packages, and conducts all necessary munging for printing tables and plots. 

```{r, descriptives.post, message=FALSE, warning=FALSE, tidy=TRUE}
# easy item descriptive statistics from the 'ltm' package
pre.items.descriptives <- descript(items[-1], chi.squared = TRUE, 
                                   B = 1000)
# extract the proportions in each categoty
pre.per <- do.call(rbind, lapply((pre.items.descriptives[2]), data.frame, 
                                 stringsAsFactors=FALSE)) %>%
  mutate(item = colnames(items)[-1]) %>%
         rename(Wrong = X0, Correct = X1) %>%
  dplyr::select(item, Wrong, Correct)

# convert to long for plotting 
pre.per.long <- gather(pre.per, cat, value, -item) %>%
  arrange(item)
```

### Analysis of mising data

Let's look at the percent of missing responses for each item. A color bar has been added to the values in the table to compare the relative proportion missing per each item. 

```{r, descriptives.1, message=FALSE, warning=FALSE, tidy=TRUE, results='asis'}
# extract the proportions in each categoty
do.call(rbind, lapply((pre.items.descriptives[7]), data.frame, 
                                 stringsAsFactors=FALSE)) %>%
  rownames_to_column("Statistic") %>%
  filter(Statistic=="missin.(%)") %>%
  gather(item, value, -Statistic) %>% 
  dplyr::select(item, value) %>%
  rename(Percent = value, Item = item) %>% 
  mutate("Percent Missing" = color_bar("lightgreen")((Percent/100)*100)) %>%
  dplyr::select(Item, "Percent Missing") %>%
  kable(digits = 2, format="html", caption="Category Utilization for pre-Administration 
        Period", escape = F) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed")) 
```

### Proportion Correct 

Let's look at the table of the proportions of rating scale category utilization to see if anything looks aberrant.

```{r, descriptives.2, message=FALSE, warning=FALSE, tidy=TRUE, results='asis'}
# print the table
pre.per %>%
  rename(Item = item) %>%
  kable(digits = 3, format="html", caption="Category Utilization for PRe-Administration 
        Period") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

A visualization is provided for another perspective to examine proportion correct.

```{r, descriptives.3, message=FALSE, warning=FALSE, tidy=TRUE, results='asis'}
# plot the proportions
p_pl1_prop <- ggplot() + geom_bar(aes(y = value, x = item, fill = cat), 
                                  data = pre.per.long, stat="identity") +
  ggtitle("Proportion Correct of SEP Items") + 
  scale_fill_ptol() + theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) 
p_pl1_prop
```

### Distribution of total score
A histogram of the total score is provided to examine whether the total score for the measure is normally distributed with no obvious ceiling or floor effects. 

```{r, descriptives.4, message=FALSE, warning=FALSE, tidy=TRUE, results='asis'}
total <- rowSums(items[-1])
histogram(~total, breaks=10)
```

## Fit the IRT Models

The syntax below fits both the Rasch model, and the 2-PL model for dichotomous data. Standard errors are calculated based on the sandwich covariance estimate, which was chosen to adjust for nesting in the data. Results for the best fitting model (i.e., lowest BIC), indicated that the 2-PL model should be utilized in subsequent model testing. 

```{r, mirt.1, message=FALSE, warning=FALSE, tidy=TRUE}
# define the number of cores for quicker processing 
mirtCluster(5)  

# run the Rasch model
model_1pl_rasch <- mirt(items[-1], 1, itemnames = 
                        c(colnames(items[-1])), 
                        itemtype = 'Rasch', 
                        technical = list(removeEmptyRows=TRUE, parallel=TRUE), 
                        empiricalhist = TRUE,
                        SE = TRUE, SE.type = 'sandwich',
                        verbose = FALSE)
summary(model_1pl_rasch)

# 2-PL model
model_2pl <- mirt(items[-1], 1, itemnames = 
                        c(colnames(items[-1])), 
                        itemtype = '2PL', 
                        technical = list(removeEmptyRows=TRUE, parallel=TRUE, NCYCLES = 10000), 
                        empiricalhist = TRUE,
                        SE = TRUE, SE.type = 'sandwich',
                        verbose = FALSE)
model_2pl


# 3-PL model (computationally singular. Sample is small, so we don't worry too much)
model_3pl <- mirt(items[-1], 1, itemnames =
                        c(colnames(items[-1])),
                        itemtype = '3PL',
                        technical = list(removeEmptyRows=TRUE, parallel=TRUE, NCYCLES = 100000),
                        empiricalhist = TRUE,
                        SE = TRUE, SE.type = 'sandwich',
                        verbose = FALSE)
model_3pl


# test the fit of 1 model vs. the other w/ BIC Table
tibble(
  "Rasch" = model_1pl_rasch@Fit$BIC,
  "2PL" = model_2pl@Fit$BIC,
  "3PL" = model_2pl@Fit$BIC)%>%
  gather(key, BIC)%>%
  arrange(BIC)

# if needed, log-likelihood test can be provided with the command below
#anova(model_1pl_rasch, model_2pl)
```

### IRT coefficients

Inspect the best model using coef(), plotting functions and goodness-of-fit functions (see the according slide and cheat sheet). 

```{r, mirt.a, message=FALSE, warning=FALSE, tidy=TRUE, results='asis'}
as.data.frame(coef(model_2pl, IRTparms = T, simplify = TRUE)) %>%
  rename(discrimination = items.a1,
         difficulty = items.d) %>%
  mutate(Item = colnames(items)[-1]) %>%
  dplyr::select(Item, discrimination, difficulty) %>%
  #arrange(-discrimination) %>%
  kable(digits = 2, format="html", caption="Item IRT Parameters", escape = F) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed")) 

#mirt:::mirt2traditional(test)
#test <- coef(model_2pl, IRTparms = T, simplify = TRUE)
```

### Category response curves

Next, the category response curves are examined. We're looking for two things: 1) the location of the item along the ability scale (i.e., difficulty), and 2) how well an item can differentiate among examinees who have abilities above and below the item location (i.e., the discrimination parameter). The steeper the curve, the better it can discriminate. Flatter curves indicate an almost equal probability of getting an item correct at either end of the ability continuum. The plots below indicate several problematic items: 2, 3, 4, 5, 7, 8, and **12**. Of these problematic items, item 12 should most certainly be removed from the analyses, as it has a negative discrimination parameter, which yields a monotonically decreasing item response function. This result indicates that people with high ability have a lower probability of responding correctly than people of low ability. The best discriminating items are: 1, 2, 10, 11, and 13, as these all exhibit the steepest curves. 
```{r, mirt.crc, message=FALSE, warning=FALSE, tidy=TRUE, results='asis'}
# table for reference
item.ref %>%
   kable(digits = 2, format="html", caption="Item Labels and Reference Numbers", escape = F) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed")) 

# create fucntion to plot combined CRC and inforamtion 
plotCRC<-function(i){
itemplot(model_2pl, i, type = 'infotrace')
}

# plot all items using the function
lapply(colnames(items)[-1], plotCRC)

plot(model_2pl, type = 'trace', auto.key = FALSE)

# plot the all of the curves in the same lattice plot. 
plot(model_2pl, type = 'trace', which.items = 1:12, facet_items=FALSE)

#
residuals(model_2pl)
```

### Model fit 

**Global Fit**: How well do these models fit the data, and do the items appear to behave well given the selected itemtypes? The `M2()` function is used to assess global model fit. Overall, the model fits the data well. 

Criteria for evaluating overall model fit:

- RMSEA <= .05

- SRMR <= .05

- Non-significant M2 statistic

- TLI and CFI >= .90

```{r, message=FALSE, warning=FALSE, tidy=TRUE}
# Global fit ---

# M2
M2_2PL <- M2(model_2pl, impute = 100, residmat = TRUE)
M2_2PL
```

**Item Fit**: The `itemfit()` in a piece-wise assessment of item fit using [Orlando and Thissen's (2000) S_X2 statistic](https://www.researchgate.net/profile/Maria_Edelen/publication/236884113_Likelihood-Based_Item-Fit_Indices_for_Dichotomous_Item_Response_Theory_Models/links/5408807e0cf23d9765b37482.pdf). An alpha of <= .01 is typically used to indicate misfitting items. Results hint at some issues with potentially misfitting items (i.e., 10, 12, and 13), but none of them reach the alpha value. Followup analyses could be conducted using [`itemGAM()`](https://www.rdocumentation.org/packages/mirt/versions/1.25/topics/itemGAM)to estimate the response curves for these items to hypothesize as to why these items are on the fringe of misfitting. 

```{r, message=FALSE, warning=FALSE, tidy=TRUE}
# Piecewise misfit ---

# item fit statistics 
itemfit_2pl <- itemfit(model_2pl, impute = 100) 

# apply a false discovery rate to the p-value 
# p.fdr <- p.adjust(itemfit_2pl$p.S_X2,"BH")
# itemfit_2pl <- cbind(itemfit_2pl, p.fdr) # bind to postvious work

# sort the item fit statistics by p-value
itemfit_2pl %>%
  slice(1:14) %>% 
  arrange(p.S_X2) %>%
  kable(digits = 2, format="html", caption="Item Fit Statistics", escape = F) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed")) 
```

### Assumptions

**Local dependence**: [Yen’s index of local dependence Q3](https://conservancy.umn.edu/bitstream/handle/11299/107543/v08n2p125.pdf?sequence=1&isAllowed=y) is provided. Q3 is a pairwise index of correlation of the residuals from the IRT model. If some sets of items present a signicant level of residual correlation, then those items can be considered as locally dependent (Yen, 1993). Q3 statistics >= .2 are automatic. The Q3 statistic index indicated some local depenence between item 13 and item 1 and item 10. Recall from the analysis of item fit that items 10 and 13 were potentially problematic. In terms of the IRT estimates this could present some issues. Some results studies have shown that if negative local dependence is not modeled, the discrimination parameters of the interdependent items are underestimated. Research also shows that the discrimination parameter (aj) depends on the difficulty of the item it interacts with (i.e., is locally dependent with), but not on the difficulty of the item itself. There are several ways to deal with this dependency: 1) model it as a testlet, incorporate it as a paramter into the model, all of which are quite time consuming and warrant their own set of analyses. A further complication with this measure is that only 2 items exhibit local dependence. To model this in `mirt` would require the analyst to specify a separate factor for the locally dependent items, in this case because there are only two locally dependent items, the slopes for these would need to be constrained equal. A seperate theta score for the conditionally dependent items would be generated, and because there are only two items you'd likely observe this score to behave eratically. If elimination of the items is not possible, then you just need to accept that the discrimnination parameters for these two items are underestimated, and not take too much stock in the difficulty estimates (i.e., if you observe these two items represent either the floor or the ceiling of the measure I'd worry about including them in the analysis). 

Example of the constrained model statement: 

  mod.statement <- '
  THETA = s_preSEP_01, s_preSEP_02, s_preSEP_03, s_preSEP_04, s_preSEP_05, s_preSEP_06, s_preSEP_07, s_preSEP_08,   s_preSEP_09, s_preSEP_11, s_preSEP_12, s_preSEP_14
  
  # two items exhibiting local dependence (should these two items not be included in the THETA call above?)
  RESID.THETA = s_preSEP_10, s_preSEP_13         
  
  # constrain the two slopes to identify the construct for the factor with local dependence
  CONSTRAIN = (s_preSEP_10, s_preSEP_13, a2)   
  COV = THETA*THETA
  '

```{r, message=FALSE, warning=FALSE, tidy=TRUE}
tidy(residuals(model_2pl, type = 'Q3', method = 'ML', suppress = .19))
```

**Unidimensionality**: A CFA is carried out to test the assumption that the measure is unidimensional. Fit statistics for the measure look OK, except the parameter for estimate for factor loading for item 12 is negative. This item was problematic in the IRT analyses above, as it displayed a negative discrimination parameter. It is an obvious candidate to remove from follow-up analyses. 
```{r, message=FALSE, warning=FALSE, tidy=TRUE, results='asis'}
model.1 <- '
  # measurement model
    factor =~ s_preSEP_01 + s_preSEP_02 + s_preSEP_03 + 
                s_preSEP_04 + s_preSEP_05 + s_preSEP_06 + 
                s_preSEP_07 + s_preSEP_08 + s_preSEP_09 + 
                s_preSEP_10 + s_preSEP_11 + s_preSEP_12 +
                s_preSEP_13 + s_preSEP_14
'
fit.1 <- cfa(model.1, data=items, std.lv = TRUE,
             ordered = c("s_preSEP_01", "s_preSEP_02", "s_preSEP_03", 
                "s_preSEP_04", "s_preSEP_05", "s_preSEP_06", 
                "s_preSEP_07", "s_preSEP_08", "s_preSEP_09", 
                "s_preSEP_10", "s_preSEP_11", "s_preSEP_12", 
                "s_preSEP_13", "s_preSEP_14"))
summary(fit.1, standardized=TRUE, fit.measures = TRUE)
```

Results testing a 1-factor solution for the measure, with item 12 removed, also fit good.
```{r, message=FALSE, warning=FALSE, tidy=TRUE, results='asis'}
# dropping item 12
model.2 <- '
  # measurement model
    factor =~ s_preSEP_01 + s_preSEP_02 + s_preSEP_03 + 
                s_preSEP_04 + s_preSEP_05 + s_preSEP_06 + 
                s_preSEP_07 + s_preSEP_08 + s_preSEP_09 + 
                s_preSEP_10 + s_preSEP_11 + 
                s_preSEP_13 + s_preSEP_14
'
fit.2 <- cfa(model.2, data=items, std.lv = TRUE,
              ordered = c("s_preSEP_01", "s_preSEP_02", "s_preSEP_03", 
                "s_preSEP_04", "s_preSEP_05", "s_preSEP_06", 
                "s_preSEP_07", "s_preSEP_08", "s_preSEP_09", 
                "s_preSEP_10", "s_preSEP_11", "s_preSEP_12", 
                "s_preSEP_13", "s_preSEP_14"))
summary(fit.2, standardized=TRUE, fit.measures = TRUE)
```

## Follow-up IRT removing item 12

```{r, message=FALSE, warning=FALSE, tidy=TRUE, results='asis'}




```

### Test Information

Examine information, SEM, and reliability for the whole measure. The plots below show that information maxmimizes around an average ability level (i.e., in the range of -2 to + 2), standard errors are lower in this range, and reliability is maxmized. It is notable that reliability within this region fail to meet the conventional critieria of <=.70. 

```{r, mirt.4, message=FALSE, warning=FALSE, tidy=TRUE, results='asis'}
# examine test information
info_model_2pl <- tibble(
  theta = seq(-6,6,.01),
  information = testinfo(model_2pl, theta),
  error = 1/sqrt(information),
  reliability = information/(information+1))

# plot test information
plot(model_2pl, type='info', MI=1000)

# plot SEM
plot(model_2pl, type='SE', MI=1000)

# plot alpha at theta levels
plot(model_2pl, type='rxx', MI=1000)
```

### Information Curves

Next, the information curves for each item are examined, looking for overlap in category curves, as well as plateaus in the information curves. Results indicate a most items provide information about students with average ability. Ideally, these information curves should show peaks that are more spread out across the range of the underlying continuum. For example, only items 2, 4, 7, and 8 provide information for students in the above average portion of the latent continuum, though for the same approximate range. This could indicate a lot of redundancy in how the item pool. Finally, the scale of the y-axis should be considered in interpretting these plots. 

```{r, message=FALSE, warning=FALSE, tidy=TRUE, results='asis'}
# table for reference
item.ref %>%
   kable(digits = 2, format="html", caption="Item Labels and Reference Numbers", escape = F) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed")) 

# create fucntion to plot combined CRC and inforamtion 
plotInfo<-function(i){
itemplot(model_2pl, i, type = 'info')
}

# plot all items using the function
lapply(colnames(items)[-1], plotInfo)
```

### Factor scores vs Standardized total scores 

The plots below indicate the association between the standardized raw scores for the measure and the several different IRT generated scores (i.e., EAP, WLE, and ML). The plots show a slight curvilinear relationship, with IRT scores for extremely low scores higher than the standardized total scores.  

```{r, message=FALSE, warning=FALSE, tidy=TRUE, results='asis'}
# Factor scores vs Standardized total scores 
fs.eap <- as.vector(fscores(model_2pl, method = "EAP", full.scores = TRUE)) 
fs.wle <- as.vector(fscores(model_2pl, method = "WLE", full.scores = TRUE)) 
fs.ml <- as.vector(fscores(model_2pl, method = "ML", full.scores = TRUE)) 
sts <- as.vector(scale(apply((items)[-1], 1, sum))) 
fscore_plot_dat <- as_data_frame(cbind(fs.eap, fs.wle, fs.ml, sts))

# IRT EAP scores vs. standardized scores
p1.eap <- ggplot(fscore_plot_dat, aes(x=sts, y=fs.eap)) + 
  geom_point()+
  geom_smooth() + 
  theme_minimal() + 
  ggtitle("IRT scores vs. Standardized Scores") + 
  labs(y="EAP IRT Score", x="Standardized Scores")

# IRT WLE scores vs. standardized scores
p1.wle <- ggplot(fscore_plot_dat, aes(x=sts, y=fs.wle)) + 
  geom_point()+
  geom_smooth() + 
  theme_minimal() + 
  ggtitle("IRT scores vs. Standardized Scores") + 
  labs(y="WLE IRT Score", x="Standardized Scores")

# IRT ML scores vs. standardized scores
p1.ml <- ggplot(fscore_plot_dat, aes(x=sts, y=fs.ml)) + 
  geom_point()+
  geom_smooth() + 
  theme_minimal() + 
  ggtitle("IRT scores vs. Standardized Scores") + 
  labs(y="ML IRT Score", x="Standardized Scores")

# plot in grid
grid.arrange( p1.eap, p1.wle, p1.ml, ncol=2)

# histogram of theta
ggplot(fscore_plot_dat, aes(fscore_plot_dat$fs)) + geom_histogram(bins=25)
```








