---
title: "IRT Analysis for the SEP Items (Pre-Administration Period)"
author: "Brandon Foster, Ph.D."
date: "`r format(Sys.time(), '%d %B, %Y, %H:%M')`"
output:
  html_document:
    theme: simplex
    toc: TRUE
    code_folding: hide
    toc_float: true
---

# Data Screeining 

First, let's bring in the data and munge it appropriately. 
```{r, message=FALSE, warning=FALSE, tidy=TRUE}
# Set working directory ----
#setwd("/Users/bfoster/Desktop/2017-edc/science-fairs-analyses")
# Load packaes ----
if (!require("pacman")) install.packages("pacman")
pacman::p_load(psych, ggplot2, ggalt, ggthemes, 
    readr, dplyr, knitr, scales, pander, kableExtra, stringr, scales,
    mirt, ltm, tidyverse, formattable, gridExtra, tidyverse, broom, lavaan)

# Import the data ----
joined.dat <- readRDS(file = "../data/joined.dat")

# Munge data ----
items <- joined.dat %>%
  dplyr::select(StudentID, s_postSEP_01, s_postSEP_02, s_postSEP_03, 
                s_postSEP_04, s_postSEP_05, s_postSEP_06, 
                s_postSEP_07, s_postSEP_08, s_postSEP_09, 
                s_postSEP_10, s_postSEP_11, s_postSEP_12, 
                s_postSEP_13, s_postSEP_14) %>%
  rename_(.dots=setNames(names(.), gsub("s_postInt_", "", names(.))))

# create dataframe for item reference
item.ref <- tibble(
  Item = colnames(items)[-1],
  Number = 1:14)
```

# Item Descriptives

The syntax below creates the item descriptive statistics using the `ltm` packages, and conducts all necessary munging for printing tables and plots. 

```{r, descriptives.post, message=FALSE, warning=FALSE, tidy=TRUE}
# easy item descriptive statistics from the 'ltm' package
post.items.descriptives <- descript(items[-1], chi.squared = TRUE, 
                                   B = 1000)
# extract the proportions in each categoty
post.per <- do.call(rbind, lapply((post.items.descriptives[2]), data.frame, 
                                 stringsAsFactors=FALSE)) %>%
  mutate(item = colnames(items)[-1]) %>%
         rename(Wrong = X0, Correct = X1) %>%
  dplyr::select(item, Wrong, Correct)

# convert to long for plotting 
post.per.long <- gather(post.per, cat, value, -item) %>%
  arrange(item)
```

## Analysis of mising data

Let's look at the percent of missing responses for each item. A color bar has been added to the values in the table to compare the relative proportion missing per each item. Nothing looks alarming. 

```{r, descriptives.1, message=FALSE, warning=FALSE, tidy=TRUE, results='asis'}
# extract the proportions in each categoty
do.call(rbind, lapply((post.items.descriptives[7]), data.frame, 
                                 stringsAsFactors=FALSE)) %>%
  rownames_to_column("Statistic") %>%
  filter(Statistic=="missin.(%)") %>%
  gather(item, value, -Statistic) %>% 
  dplyr::select(item, value) %>%
  rename(Percent = value, Item = item) %>% 
  mutate("Percent Missing" = color_bar("lightgreen")((Percent/100)*100)) %>%
  dplyr::select(Item, "Percent Missing") %>%
  kable(digits = 2, format="html", caption="Category Utilization for post-Administration 
        Period", escape = F) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed")) 
```

## Proportion Correct 

Let's look at the table of the proportions correct for each item in order to see if anything looks aberrant. 

```{r, descriptives.2, message=FALSE, warning=FALSE, tidy=TRUE, results='asis'}
# print the table
post.per %>%
  rename(Item = item) %>%
  kable(digits = 3, format="html", caption="Category Utilization for post-Administration 
        Period") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

A visualization is provided for another perspective to examine proportion correct. Note, that I've added a horizontal line at .50 for reference. 

```{r, descriptives.3, message=FALSE, warning=FALSE, tidy=TRUE, results='asis'}
# plot the proportions
p_pl1_prop <- ggplot() + geom_bar(aes(y = value, x = item, fill = cat), 
                                  data = post.per.long, stat="identity") +
  ggtitle("Proportion Correct of SEP Items") + 
  labs(x="Items", y="Proportion of Response Option", fill="Category") +
  scale_fill_ptol() + theme_minimal() +
  geom_hline(yintercept = .5) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) 
p_pl1_prop
```

## Distribution of total score
A histogram of the total score is provided to examine whether the total score for the measure is normally distributed with no obvious ceiling or floor effects. 

```{r, descriptives.4, message=FALSE, warning=FALSE, tidy=TRUE, results='asis'}
total <- rowSums(items[-1])
histogram(~total, breaks=10)
```

# Fit the IRT Models

The syntax below fits both the Rasch model, and the 2 & 3-PL models for dichotomous data. Standard errors are calculated based on the sandwich covariance estimate, which was chosen to adjust for nesting in the data. Results for the best fitting model (i.e., lowest sample adjusted BIC), indicated that the 2-PL model should be utilized in subsequent model testing. 

```{r, mirt.1, message=FALSE, warning=FALSE, tidy=TRUE}
# define the number of cores for quicker processing 
mirtCluster(5)  

# drop missing
items.noNA <- na.omit(items)

# run the Rasch model
set.seed(8675309)
model_1pl_rasch <- mirt(items.noNA[-1], 1, 
                        itemtype = 'Rasch', 
                        technical = list(removeEmptyRows=TRUE, parallel=TRUE, NCYCLES = 8000), 
                        SE = TRUE, SE.type = 'sandwich',
                        verbose = FALSE)

#check model convergence
model_1pl_rasch 

# 2-PL model
set.seed(8675309)
model_2pl <- mirt(items.noNA[-1], 1, 
                        itemtype = '2PL', 
                        technical = list(removeEmptyRows=TRUE, parallel=TRUE, NCYCLES = 8000), 
                        SE = TRUE, SE.type = 'sandwich',
                        verbose = FALSE)
# check to see that model converged
model_2pl

# 3-PL model (computationally singular. Sample is small, so we don't worry too much)
set.seed(8675309)
model_3pl <- mirt(items.noNA[-1], 1, 
                        itemtype = '3PL',
                        technical = list(removeEmptyRows=TRUE, parallel=TRUE, NCYCLES = 80000),
                        SE = TRUE, SE.type = 'sandwich',
                        verbose = FALSE)
# check model convergence
model_3pl

# examine sample adjusted BIC for best fitting model
mods.SABIC <- tibble(
  "Rasch" = model_1pl_rasch@Fit$SABIC,
  "2PL" = model_2pl@Fit$SABIC,
  "3PL" = model_3pl@Fit$SABIC)%>%
  gather(key, SABIC)%>%
  arrange(SABIC)

# print table
mods.SABIC %>%
kable(digits = 2, format="html", caption="Model Selection With SABIC", escape = F) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed")) 

# if needed, log-likelihood test can be provided with the command below
#anova(model_1pl_rasch, model_2pl)
```

## IRT coefficients

Inspect the parameter (i.e., IRT adjusted), for the best fitting model. Again, item 12 has a negative discrimination. 

```{r, mirt.a, message=FALSE, warning=FALSE, tidy=TRUE, results='asis'}
as.data.frame(coef(model_2pl, IRTparms = T, simplify = TRUE)) %>%
  rename(discrimination = items.a1,
         difficulty = items.d) %>%
  mutate(Item = colnames(items)[-1]) %>%
  dplyr::select(Item, discrimination, difficulty) %>%
  #arrange(-discrimination) %>%
  kable(digits = 2, format="html", caption="Item IRT Parameters", escape = F) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed")) 

#mirt:::mirt2traditional(model_2pl)
```

## Category response curves

Next, the category response curves are examined. We're looking for two things: 1) the location of the item along the ability scale (i.e., difficulty), and 2) how well an item can differentiate among examinees who have abilities above and below the item location (i.e., the discrimination parameter). The steeper the curve, the better it can discriminate. Flatter curves indicate an almost equal probability of getting an item correct at either end of the ability continuum. The plots below indicate several problematic items: 2, 3, 4, 5, 7, 8, and **12**. Of these problematic items, item 12 should most certainly be removed from the analyses, as it has a negative discrimination parameter, which yields a monotonically decreasing item response function. This result indicates that people with high ability have a lower probability of responding correctly than people of low ability. The best discriminating items are: 1, 2, 10, 11, and 13, as these all exhibit the steepest curves. 

Reference table for CRC plots: 
```{r, message=FALSE, warning=FALSE, tidy=TRUE, tidy=TRUE}
# table for reference
item.ref %>%
   kable(digits = 2, format="html", caption="Item Labels and Reference Numbers", escape = F) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed")) 
```

Generate CRCs for each item:
```{r, message=FALSE, warning=FALSE, tidy=TRUE, tidy=TRUE}
# create function to plot combined CRC and inforamtion 
plotCRC<-function(i){
itemplot(model_2pl, i, type = 'infotrace')
}

# plot all items using the function
lapply(colnames(items)[-1], plotCRC)
```

Plot CRCs in same frame so that difficulty of items can be visually examined:
```{r, message=FALSE, warning=FALSE, tidy=TRUE, tidy=TRUE}
# plot the all of the curves in the same lattice plot. 
plot(model_2pl, type = 'trace', which.items = 1:14, facet_items=FALSE)
```

## Model fit 

**Global Fit**: How well do these models fit the data, and do the items appear to behave well given the selected itemtypes? The `M2()` function is used to assess global model fit. Overall, the model fits the data well.

Criteria for evaluating overall model fit:

- RMSEA <= .05

- SRMR <= .05

- Non-significant M2 statistic

- TLI and CFI >= .90

```{r, message=FALSE, warning=FALSE, tidy=TRUE}
# Global fit ---
# M2
M2_model_2pl <- M2(model_2pl, impute = 100, residmat = FALSE)

M2_model_2pl %>%
kable(digits = 3, format="html", caption="Global Fit Statistics", escape = F) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed")) 
```

**Item Fit**: The `itemfit()` in a piece-wise assessment of item fit using [Orlando and Thissen's (2000) S_X2 statistic](https://www.researchgate.net/profile/Maria_Edelen/publication/236884113_Likelihood-Based_Item-Fit_Indices_for_Dichotomous_Item_Response_Theory_Models/links/5408807e0cf23d9765b37482.pdf). An alpha of <= .01 is typically used to indicate misfitting items. Results hint at some issues with potentially misfitting items (i.e., 12, 10, 5, and 8), but none of them reach the critical alpha value. Followup analyses could be conducted using [`itemGAM()`](https://www.rdocumentation.org/packages/mirt/versions/1.25/topics/itemGAM)to estimate the response curves for these items to hypothesize as to why these items are on the fringe of misfitting. 

```{r, message=FALSE, warning=FALSE, tidy=TRUE}
# Piecewise misfit ---

# item fit statistics 
itemfit_2pl <- itemfit(model_2pl, impute = 100) 

# apply a false discovery rate to the p-value 
# p.fdr <- p.adjust(itemfit_2pl$p.S_X2,"BH")
# itemfit_2pl <- cbind(itemfit_2pl, p.fdr) # bind to postvious work

# sort the item fit statistics by p-value
itemfit_2pl %>%
  slice(1:14) %>% 
  arrange(p.S_X2) %>%
  kable(digits = 2, format="html", caption="Item Fit Statistics", escape = F) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed")) 

# item GAM ----
# items.na <- na.omit(items)
# model_2pl.na <- mirt(items.na[-1], 1, 
#                         itemtype = '2PL', 
#                         technical = list(removeEmptyRows=TRUE, parallel=TRUE, NCYCLES = 8000), 
#                         SE = TRUE, SE.type = 'sandwich',
#                         verbose = FALSE)
# colnames(items.na)
# Theta <- fscores(model_2pl.na, method = "MAP", full.scores = TRUE)
# IG12 <- itemGAM(items.na[,13], Theta) 
# summary(IG12)
# plot(IG12)
```

## Assumptions

**Local dependence**: [Yen’s index of local dependence Q3](https://conservancy.umn.edu/bitstream/handle/11299/107543/v08n2p125.pdf?sequence=1&isAllowed=y) is provided. Q3 is a pairwise index of correlation of the residuals from the IRT model. If some sets of items present a significant level of residual correlation, then those items can be considered as locally dependent (Yen, 1993). Q3 statistics >= .2 are automatic. The Q3 statistic index indicated some local dependence between item 13 and item 1 and item 10. Recall from the analysis of item fit that items 10 and 13 were potentially problematic. Several items exhibited issues with local dependence: 7, 9, 10, and 13. 

Example of the constrained model statement: 
```{r, message=FALSE, warning=FALSE, tidy=TRUE}
  mod.statement <- '
  THETA = s_preSEP_01, s_preSEP_02, s_preSEP_03, s_preSEP_04, s_preSEP_05, s_preSEP_06, s_preSEP_07,
  s_preSEP_08,   s_preSEP_09, s_preSEP_10, s_preSEP_11, s_preSEP_12, s_preSEP_13, s_preSEP_14
  
  # two items exhibiting local dependence (should these two items not be included in the THETA call above?)
  RESID.THETA = s_preSEP_7, s_preSEP_09,  s_preSEP_10, s_preSEP_13         
  
  # constrain the two slopes to identify the construct for the factor with local dependence
  CONSTRAIN = (s_preSEP_7, s_preSEP_09,  s_preSEP_10, s_preSEP_13 , a2)   
  COV = THETA*THETA #note THETA 1 and THETA 2 are orthogonal
  '
```

```{r, message=FALSE, warning=FALSE, tidy=TRUE}
tidy(residuals(model_2pl, type = 'Q3', method = 'ML', suppress = .19))
```

**Unidimensionality**: A CFA is carried out to test the assumption that the measure is unidimensional. Fit statistics for the measure look OK, except the parameter for estimate for factor loading for item 12 is negative. This item was problematic in the IRT analyses above, as it displayed a negative discrimination parameter. It is an obvious candidate to remove from follow-up analyses. 

```{r, message=FALSE, warning=FALSE, tidy=TRUE}
model.1 <- '
  # measurement model
    factor =~ s_postSEP_01 + s_postSEP_02 + s_postSEP_03 + 
                s_postSEP_04 + s_postSEP_05 + s_postSEP_06 + 
                s_postSEP_07 + s_postSEP_08 + s_postSEP_09 + 
                s_postSEP_10 + s_postSEP_11 + s_postSEP_12 +
                s_postSEP_13 + s_postSEP_14
'
fit.1 <- cfa(model.1, data=items.noNA, std.lv = TRUE,
             ordered = c("s_postSEP_01", "s_postSEP_02", "s_postSEP_03", 
                "s_postSEP_04", "s_postSEP_05", "s_postSEP_06", 
                "s_postSEP_07", "s_postSEP_08", "s_postSEP_09", 
                "s_postSEP_10", "s_postSEP_11", "s_postSEP_12", 
                "s_postSEP_13", "s_postSEP_14"))
fitMeasures(fit.1)
summary(fit.1, fit.measures=TRUE)
```

Results testing a 1-factor solution for the measure, with item 12 removed, also look good.
```{r, message=FALSE, warning=FALSE, tidy=TRUE}
# dropping item 12
model.2 <- '
  # measurement model
    factor =~ s_postSEP_01 + s_postSEP_02 + s_postSEP_03 + 
                s_postSEP_04 + s_postSEP_05 + s_postSEP_06 + 
                s_postSEP_07 + s_postSEP_08 + s_postSEP_09 + 
                s_postSEP_10 + s_postSEP_11 + 
                s_postSEP_13 + s_postSEP_14
'
fit.2 <- cfa(model.2, data=items.noNA, std.lv = TRUE,
              ordered = c("s_postSEP_01", "s_postSEP_02", "s_postSEP_03", 
                "s_postSEP_04", "s_postSEP_05", "s_postSEP_06", 
                "s_postSEP_07", "s_postSEP_08", "s_postSEP_09", 
                "s_postSEP_10", "s_postSEP_11", "s_postSEP_12", 
                "s_postSEP_13", "s_postSEP_14"))
fitMeasures(fit.2)
```

## Follow-up IRT removing item 12
```{r, message=FALSE, warning=FALSE, tidy=TRUE, results='asis'}
# 2-PL model
set.seed(8675309)
#colnames(items)
items.noNA_no12 <- dplyr::select(items.noNA, StudentID, s_postSEP_01, s_postSEP_02, s_postSEP_03, s_postSEP_04, s_postSEP_05, s_postSEP_06,
                     s_postSEP_07, s_postSEP_08, s_postSEP_09, s_postSEP_10, s_postSEP_11, s_postSEP_13, s_postSEP_14)
model_2pl_b <- mirt(items.noNA_no12[-1], 1, 
                        itemtype = '2PL', 
                        technical = list(removeEmptyRows=TRUE, parallel=TRUE, NCYCLES = 8000), 
                        SE = TRUE, SE.type = 'sandwich',
                        verbose = FALSE)
# check to see that model converged
model_2pl_b

# item parameter statistics ---- 
as.data.frame(coef(model_2pl_b, IRTparms = T, simplify = TRUE)) %>%
  rename(discrimination = items.a1,
         difficulty = items.d) %>%
  mutate(Item = colnames(items.noNA_no12)[-1]) %>%
  dplyr::select(Item, discrimination, difficulty) %>%
  #arrange(-discrimination) %>%
  kable(digits = 2, format="html", caption="Item IRT Parameters", escape = F) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed")) 

# item fit statistics ---- 
itemfit_2pl_b <- itemfit(model_2pl_b, impute = 100) 

# apply a false discovery rate to the p-value 
# p.fdr <- p.adjust(itemfit_2pl$p.S_X2,"BH")
# itemfit_2pl <- cbind(itemfit_2pl, p.fdr) # bind to postvious work

# sort the item fit statistics by p-value
itemfit_2pl_b %>%
  slice(1:13) %>% 
  arrange(p.S_X2) %>%
  kable(digits = 2, format="html", caption="Item Fit Statistics", escape = F) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# plot the all of the curves in the same lattice plot ----
plot(model_2pl_b, type = 'trace', which.items = 1:13, facet_items=FALSE)
```

## Test Information

Examine information, SEM, and reliability for the whole measure. The plots below show that information maximizes around an average ability level (i.e., in the range of -2 to + 2), standard errors are lower in this range, and reliability is maximized. 

```{r, message=FALSE, warning=FALSE, tidy=TRUE, results='asis'}
# examine test information
info_model_2plb <- tibble(
  theta = seq(-6,6,.01),
  information = testinfo(model_2pl_b, theta),
  error = 1/sqrt(information),
  reliability = information/(information+1))

# plot test information
plot(model_2pl_b, type='info', MI=1000)

# plot SEM
plot(model_2pl_b, type='SE', MI=1000)

# plot alpha at theta levels
plot(model_2pl_b, type='rxx', MI=1000)
```

## Information Curves

Next, the information curves for each item are examined, looking for overlap in category curves, as well as plateaus in the information curves. Results indicate a most items provide information about students with average ability. Ideally, these information curves should show peaks that are more spread out across the range of the underlying continuum. For example, only items 2, 4, 7, and 8 provide information for students in the above average portion of the latent continuum, though for the same approximate range. This could indicate a lot of redundancy in how the item pool. Finally, the scale of the y-axis should be considered in interpreting these plots. 

```{r, message=FALSE, warning=FALSE, tidy=TRUE, results='asis'}
# table for reference
item.ref %>%
   kable(digits = 2, format="html", caption="Item Labels and Reference Numbers", escape = F) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed")) 
```

Plot the information curves in one grid for easy comparison. 
```{r, message=FALSE, warning=FALSE, tidy=TRUE, results='asis'}
plot(model_2pl, type = 'infotrace', which.items = 1:14, facet_items=FALSE)
```

## Factor scores vs Standardized total scores 

The plots below indicate the association between the standardized raw scores for the measure and the several different IRT generated MAP scores. 

```{r, message=FALSE, warning=FALSE, tidy=TRUE, results='asis'}
# Factor scores vs Standardized total scores 
#fs.map <- as.vector(
Theta_postSep <-  fscores(model_2pl_b, method = "MAP", full.scores = TRUE)
STS_postSep <- as.vector(scale(apply((items.noNA_no12)[-1], 1, sum))) 
TOTAL_postSep<- apply((items.noNA_no12)[-1], 1, sum)

# save the factor scores
post.sep.theta <- cbind(
  items.noNA_no12[1], Theta_postSep, TOTAL_postSep, STS_postSep
) %>%
  rename(Theta_postSep = F1)

# write the data
write_csv(post.sep.theta, "../data/post.sep.theta.csv")

# IRT MAP scores vs. standardized scores
p1.map <- ggplot(post.sep.theta, aes(x=TOTAL_postSep, y=Theta_postSep)) + 
  geom_point()+
  geom_smooth() + 
  theme_minimal() + 
  ggtitle("IRT scores vs. Standardized Scores") + 
  labs(y="MAP IRT Score", x="Standardized Scores")

p1.map

# histogram of theta
ggplot(post.sep.theta, aes(post.sep.theta$Theta_postSep)) + geom_histogram(bins=25)

# potential outlier: 1256121
```
